A Large and Variable Leading Tail of Helium in a Hot Saturn Undergoing Runaway Inflation

### 2023-07-13 paper/paper1/ms.tex
Submitted version of the paper!
Ok, this is the revised submission with the shorter abstract!

% and more...

\begin{abstract}
    Atmospheric escape shapes the fate of exoplanets, with statistical evidence for transformative mass loss imprinted across the mass-radius-insolation distribution. The Habitable Zone Planet Finder (HPF) Spectrograph Exospheres program aims to expand the search across a diverse range of planet and host star properties via the 10833 \AA~ Helium triplet, a sensitive probe of active atmospheric escape. Here we present a detection of up to 10\% absorption depth during transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67~b. The 13.8 hours of on-sky integration time over 39 nights sample the entire planet orbit, uncovering excess Helium absorption preceding the transit by up to 130 planetary radii in a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe and advecting most of the gas into the stellar—and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry.  We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary winds.  We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star.  The large mass loss in HAT-P-67~b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly.  We contrast two physical mechanisms for runaway evaporation: Ohmic dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}

>>>>

% This is the submitted version: July 13, 2023 at 3:13 pm central 
% Revised at 3:31 pm to have a shorter abstract

\begin{abstract}
    Atmospheric escape shapes the fate of exoplanets, with statistical evidence for transformative mass loss imprinted across the mass-radius-insolation distribution. Here we present transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67~b.  The Habitable Zone Planet Finder (HPF) spectra show a detection of up to 10\% absorption depth of the 10833 \AA~ Helium triplet. The 13.8 hours of on-sky integration time over 39 nights sample the entire planet orbit, uncovering excess Helium absorption preceding the transit by up to 130 planetary radii in a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe and advecting most of the gas into the stellar---and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry. We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary winds. We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star. The large mass loss in HAT-P-67 b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly. We contrast two physical mechanisms for runaway evaporation: Ohmic dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}
